---
title: "Wright-Fisher Simulation with Selection and Drift"
output: html_document
date: "2026-02-06"
author: "Jinliang Yang"
coding_tool: "Codex"
model: "GPT-5"
reviewer: "Jinliang Yang"
review_status: "Human reviewed"
---

# Wright-Fisher simulation with selection and drift

## Purpose

This document defines a reusable Wright-Fisher simulation function for allele-frequency change under selection and drift in a diploid population.

The main goal is to keep the method readable, reproducible, and easy to adapt for later simulation studies. The code below is written as a template: first explain what the code is doing, then define parameters, and only then provide the implementation.

## Relative path setup

This setup chunk makes the document run relative to the project root instead of depending on the current working directory in the R session.

```{r setup, include=TRUE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE
)

current_input_dir <- normalizePath(
  dirname(knitr::current_input(dir = TRUE)),
  winslash = "/",
  mustWork = TRUE
)

project_root <- normalizePath(
  file.path(current_input_dir, ".."),
  winslash = "/",
  mustWork = TRUE
)

knitr::opts_knit$set(root.dir = project_root)
```

## Method overview

We simulate allele-frequency trajectories across generations under a Wright-Fisher model with viability selection.

At each generation, the workflow is:

1. compute genotype frequencies under Hardy-Weinberg expectations from the current allele frequency
2. apply genotype fitness values to obtain the post-selection allele frequency
3. sample the next generation by binomial drift using `2 * Ne` gene copies

This template is useful when you want a compact simulation function that is still explicit enough for human review.

## Parameters

The simulation function uses the following parameters.

- `p0`: initial allele frequency of the focal allele
- `Ne`: effective population size for a diploid population
- `s`: selection coefficient for the homozygous focal genotype relative to the reference homozygote
- `gens`: number of generations to simulate
- `h`: dominance coefficient, where `h = 0.5` gives an additive heterozygote effect

## Code

The code chunk below implements the simulation function described above.

```{r wf-sel-drift-function}
wf_sel_drift <- function(p0, Ne, s, gens = 60, h = 0.5) {
  # p0: initial allele frequency of allele A
  # Ne: effective population size in a diploid population
  # s: selection coefficient for AA relative to aa
  # gens: number of generations to simulate
  # h: dominance coefficient for Aa, where fitness is 1 + h * s

  p <- numeric(gens + 1)
  p[1] <- p0

  for (t in seq_len(gens)) {
    pt <- p[t]

    # Genotype frequencies before selection.
    pAA <- pt^2
    pAa <- 2 * pt * (1 - pt)
    paa <- (1 - pt)^2

    # Relative fitness values.
    wAA <- 1 + s
    wAa <- 1 + h * s
    waa <- 1

    # Mean fitness after weighting genotype frequencies.
    wbar <- pAA * wAA + pAa * wAa + paa * waa

    # Allele frequency after selection and before drift.
    p_sel <- (pAA * wAA + 0.5 * pAa * wAa) / wbar

    # Random genetic drift from binomial sampling of 2Ne gene copies.
    k <- rbinom(n = 1, size = 2 * Ne, prob = p_sel)
    p[t + 1] <- k / (2 * Ne)
  }

  p
}
```

## Example usage

This chunk gives a minimal example for testing the function.

```{r example-usage}
example_traj <- wf_sel_drift(
  p0 = 0.1,
  Ne = 100,
  s = 0.05,
  gens = 60,
  h = 0.5
)

head(example_traj)
```

## Notes for adaptation

When adapting this template for another analysis, keep the same order:

1. document the purpose in plain language
2. define reproducible path handling
3. explain parameters before the code chunk
4. keep a short runnable example for smoke testing
